---
title: "Read observations from CSV files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Read observations from CSV files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

_harpIO_ only includes functionality to read observations from _vobs_ files, 
though observations could come in many formats. A simple format could be a 
comma separated variables (CSV) text file. This vignette demonstrates how you 
can make a function that `read_obs_file()` or `read_obs()` could call to read 
in CSV files following a particular format. 

Let's say our CSV data have columns for station ID ("stationId"), station 
elevation ("stationElevation"), and columns for the variables 2m temperature 
("TA"), 10m wind speed ("FF10"), 1-hour precipitation ("RR1"), 2m relative 
humidity, ("RH2") and mean sea level pressure ("MSLP"), and that there is one 
file for each observation time, We'll begin by making a function to make some 
fake data.
```{r setup}
library(harpIO)
library(tibble)
```

```{r fake-data-func}
make_fake_data <- function(num_stn) {

  set.seed(num_stn) # so stationElevation is the same every time
  station_elevs <- round(runif(num_stn, 0, 350)) 
  set.seed(Sys.time())

  tibble::tibble(
    stationId        = seq_len(num_stn),
    stationElevation = station_elevs,
    TA               = round(runif(num_stn, 5, 35), 1),
    FF10             = round(runif(num_stn, 0.25, 35), 1),
    RR1              = round(round(rgamma(num_stn, 1, 1)) * runif(num_stn, 1, 2), 1),
    RH2              = round(runif(num_stn, 0.5, 1), 2),
    MSLP             = round(runif(num_stn, 970, 1030))
  )
}
```

And then another function to call `make_fake_data()` that will write those data 
to CSV files for a range of date-times. 

```{r write-fake-data}
make_csv_files <- function(dttm, num_stn, file_path) {

  invisible(lapply(
    dttm,
    \(x) {
      df <- make_fake_data(num_stn)
      file_name <- file.path(file_path, paste0("obs_", x, ".csv"))
      write.table(df, file_name, sep = ",", row.names = FALSE)
    }
  ))
}
```

And run the function to create 24 hours of fake data, saving to `tempdir()`.

```{r run-fake-data}
make_csv_files(seq_dttm(2023112800, 2023112823), 30, tempdir())
```

Now we have a bunch of CSV files we need to function that will read them in and 
return data in a format that `read_obs_file()` and `read_obs()` can use. There 
are a couple of important things to remember - the function should return a 
named list of data frames and those data frames should be one for the data 
themselves and one that contains information about the units and accumulation 
times of the observations. We will call these data frames "synop" for the data 
and "synop_params" for the information about units. "synop" is chosen so that 
harp knows that these are near surface observations akin to official SYNOP 
meteorological observations. We should also use a parameter naming convention
that harp knows (see `show_param_defs()`) as well as column names "SID" for 
station ID and "elev" for station elevation. Note that we use namespaces for all 
functions from packages to future proof the function in case we want to use it 
in a package in the future. 

```{r read-function}
read_csv_obs <- function(file_name, dttm, parameter = NULL, ...) {
  
  # read the csv data using read.csv
  obs_data <- read.csv(file_name)
  
  # add a column for the date-time
  obs_data <- dplyr::mutate(
    obs_data, valid_dttm = dttm, .before = dplyr::everything()
  )
  
  # Change column names to names harp knows about using psub()
  # We wrap in suppressWarnings as we don't want it to warn about 
  # substitutions that don't exist in case not all files contain all the 
  # expected columnn names. 
  colnames(obs_data) <- suppressWarnings(psub(
    colnames(obs_data),
    c("stationId", "stationElevation", "TA", "FF10", "RR1", "RH2", "MSLP"),
    c("SID", "elev", "T2m", "S10m", "AccPcp1h", "RH2m", "Pmsl")
  ))
  
  # Only return parameters that have been asked for
  if (length(parameter) > 0 && !all(is.na(parameter))) {
    parameter <- stats::na.omit(parameter)
    obs_data <- dplyr::select(
      obs_data, 
      dplyr::any_of(c("valid_dttm", "SID", "elev", parameter))
    )
  }
  
  # Make a data frame for parameter units
  obs_units <- tibble::tribble(
    ~parameter, ~accum_hours, ~units,
    "T2m"     , 0           , "degC",
    "S10m"    , 0           , "m/s",
    "AccPcp1h", 1           , "kg/m^2",
    "RH2m"    , 0           , "fraction",
    "Pmsl"    , 0           , "hPa",
  )
  
  # Filter the obs_units data frame to only those parameters that are in the 
  # data
  obs_units <- dplyr::filter(obs_units, .data$parameter %in% colnames(obs_data))
  
  # return the data as a named list
  list(synop = obs_data, synop_params = obs_units)
}
```

We can now test the `read_csv_obs()` function on its own or called by 
`read_obs_file()` by setting `file_format = "csv_obs"`.

```{r basic-read}
read_csv_obs(file.path(tempdir(), "obs_2023112812.csv"), 2023112812)
read_obs_file(
  file.path(tempdir(), "obs_2023112812.csv"), NULL, 
  dttm = 2023112812, file_format = "csv_obs"
)
read_obs_file(
  file.path(tempdir(), "obs_2023112812.csv"), c("T2m", "S10m"), 
  dttm = 2023112812, file_format = "csv_obs"
)
```
Note the difference in the valid_dttm column. `read_csv_obs()` returns the date 
times that were given to it, whereas `read_obs_file()` converts the 
__valid_dttm__ column to UNIX epoch time format - that is number of seconds 
since 00:00:00 1 Januray 1970. This is because this is the most efficient way 
to store and process date-time data. 

If we want to use the observations throughout harp, for example for point 
observations, it is most efficient to convert the observations to SQLite format, 
which is much quicker to access and filter to what we need using 
`read_point_obs()`. This process is done using `read_obs()` and setting 
output_format_opts. 

```{r read-obs}
read_obs(
  seq_dttm(2023112800, 2023112823), 
  NULL, 
  file_format        = "csv_obs", 
  file_path          = tempdir(),
  file_template      = "obs_{YYYY}{MM}{DD}{HH}.csv",
  output_format_opts = obstable_opts(file.path(tempdir(), "OBSTABLE"))
)
```

And then read the data back in using `read_point_obs()`

```{r read-point-obs}
read_point_obs(
  seq_dttm(2023112803, 2023112819), "T2m", 
  obs_path = file.path(tempdir(), "OBSTABLE"), stations = c(4, 6, 9)
)
```

